Using CopyKAT to idenfity tumor cells in Ewing sarcoma samples

### 2024-05-10 analyses/ewings-cell-type/03-copykat.Rmd
add common CNAs to analysis

- Visualize diploid and anueploid cells on the UMAP. 
- Calculate the confusion matrix comparing manual annotations of tumor cells to using `CopyKAT` to annotate tumor cells. 
- Compare the annotations from `CopyKAT` to cell type annotations from `SingleR` and `CellAssign`.

@@

# copykat results 
results_dir <- file.path(module_base, "results", "copykat", params$library_id)
results_filenames <- c(
  no_ref = glue::glue("{params$library_id}_no_normal_copykat_prediction.txt"), 
  with_ref = glue::glue("{params$library_id}_with_normal_copykat_prediction.txt")
) 
ck_results_files <- file.path(results_dir, results_filenames) |> 
  purrr::set_names(names(results_filenames))
```

@@

# read in ck results from both reference types (no_normal and with_normal)
ck_results_df <- ck_results_files |> 
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "reference_used")
```

@@

@@

@@

```{r}
filtered_cnv_df <- cnv_df |> 
  dplyr::filter(copykat.pred != "not.defined") |> 
caret_df_list <- filtered_cnv_df |> 
  dplyr::mutate(copykat = ifelse(
    # use str_detect for test data
    copykat.pred == "diploid", "Normal", "Tumor"
  )) |> 
  # make tumor the positive class
  dplyr::mutate(
    copykat = forcats::fct_relevel(copykat, "Tumor"),
    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
  ) |> 
  split(cnv_df$reference_used)

caret_df |> 
  purrr::imap(\(df, ref_type){
    glue::glue("\n\n================== {ref_type} ====================")
    caret::confusionMatrix(
      table(
        df$marker_gene_classification, 
        df$copykat)
    )
  })
```

@@

```{r}
# calculate Jaccard similarity index for each reference type
ref_types <- c("no_ref", "with_ref")
jaccard_matrices <- ref_types |>
  purrr::map(\(ref_type) {
    jaccard_df <- cnv_df |> 
      dplyr::filter(reference_used == ref_type)

    make_jaccard_matrix(
      cnv_df,
      "marker_gene_classification",
      "copykat.pred"
    )
  }) |> 
  purrr::set_names(ref_types)
```

@@

- `CopyKAT` seems to do a good job in this Ewing sarcoma sample of identifying aneuploid and diploid cells. 
- The annotations from `CopyKAT` mostly line up with annotations obtained using marker gene annotation. 
- Using a reference of normal cells doesn't change the overall results but does reduce the run time for `CopyKAT`.

>>>>

- Visualize diploid and anueploid cells on the UMAP. 
- Evaluate common copy number gains and losses in Ewing sarcoma. 
- Calculate the confusion matrix comparing manual annotations of tumor cells to using `CopyKAT` to annotate tumor cells. 
- Compare the annotations from `CopyKAT` to cell type annotations from `SingleR` and `CellAssign`.

@@

# copykat results 
results_dir <- file.path(module_base, "results", "copykat", params$library_id)

# predictions files 
predictions_file <- glue::glue("{params$library_id}_copykat_prediction.txt")

predictions_paths <- c(
  no_ref = file.path(results_dir, "no_normal", predictions_file),
  with_ref = file.path(results_dir, "endothelial_normal", predictions_file)
)

# full results gene by cell 
full_ck_result_file <- glue::glue("{params$library_id}_copykat_CNA_raw_results_gene_by_cell.txt")

full_ck_result_paths <- c(
  no_ref = file.path(results_dir, "no_normal", full_ck_result_file),
  with_ref = file.path(results_dir, "endothelial_normal", full_ck_result_file)
)

```

@@

# read in ck predictions from both reference types (no_normal and with_normal)
ck_results_df <- predictions_paths |> 
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "reference_used")

# read in full gene by cell copy number detection results 
full_ck_results_df <- full_ck_result_paths |> 
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "reference_used")
```

@@

Below we look at the heatmaps produced by `CopyKAT`.

### Heatmap without reference

![](`r file.path(results_dir, "no_normal", "SCPCL000822_copykat_heatmap.jpeg")`)

### Heatmap with endothelial cells as reference

![](`r file.path(results_dir, "endothelial_normal", "SCPCL000822_copykat_heatmap.jpeg")`)

Both of these heatmaps appear to show some increase in observed CNVs in aneuploid vs. diploid which is to be expected.

@@

## Validate common CNAs found in Ewing sarcoma

To validate some of these annotations, we can also look at some commonly found copy number variations found in Ewing sarcoma patients. 
There are a few known copy number variations in Ewing's sarcoma:

- Gain of Chr8
- Gain of Chr12
- Gain of Chr1p
- Loss of Chr16q

Although these are the most frequent, there are patients who do not have any of these alterations and patients that only have some of these alterations. 
See [Tirode et al.,](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4264969/) and [Crompton et al.](https://doi.org/10.1158/2159-8290.CD-13-1037).

`CopyKAT` outputs a matrix that contains the estimated copy numbers for each gene in each cell. 
We can read that in and look at the mean estimated copy numbers for each chromosome across each cell. 
We might expect that tumor cells would show an increased estimated copy number in Chr8, Chr12, and/or Chf 1 and a loss of Chr16.

```{r}
# for every cell, calculate the mean detection level across all genes in a given chromosome
full_cnv_df <- full_ck_results_df |> 
  tidyr::pivot_longer(cols = -c(
    reference_used,
    abspos, 
    chromosome_name, 
    start_position, 
    end_position, 
    ensembl_gene_id, 
    hgnc_symbol, 
    band),
                      names_to = "barcodes",
                      values_to = "cnv_detection") |> 
  dplyr::group_by(chromosome_name, barcodes, reference_used) |> 
  dplyr::summarise(mean_cnv_detection = mean(cnv_detection))

# filter to only have cnvs found in Ewings to keep things a bit smaller when joining
ewings_chrs <- c(1, 8, 12, 16)

ewings_cnv_df <- full_cnv_df |> 
  dplyr::filter(chromosome_name %in% ewings_chrs)

# join with 
cnv_df <- cnv_df |> 
  dplyr::left_join(ewings_cnv_df, by = c("barcodes", "reference_used"))
```

```{r}
# create faceted UMAPs showing estimation of CNV detection across each chr of interest 
ewings_chrs |> 
  purrr::map(\(chr){
    plot_df <- cnv_df |> 
      dplyr::filter(chromosome_name == chr)

    ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = mean_cnv_detection)) +
      geom_point(alpha = 0.5, size = 0.5) +
      theme_bw() +
      scale_color_viridis_c() +
      facet_wrap(vars(reference_used)) + 
      labs(title = glue::glue("chr{chr}"))
  })
```

Looking at the UMAPs, I don't see any really obvious differences between CNV estimation and groups of cells. 
Let's look at the distribution of CNV estimation in cells that are called aneuploid and diploid by `CopyKAT`.

```{r}
# create faceted density plots showing estimation of CNV detection across each chr of interest 
# colored by aneuploid/diploid estimation 
ewings_chrs |> 
  purrr::map(\(chr){
    plot_df <- cnv_df |> 
      dplyr::filter(chromosome_name == chr)

    ggplot(plot_df, aes(x = mean_cnv_detection, color = copykat.pred)) +
      geom_density() +
      theme_bw() +
      facet_wrap(vars(reference_used)) + 
      labs(title = glue::glue("chr{chr}"))
  })
```

It looks like there is a slight increase in chr8 and chr12 indicating maybe a copy number gain for those chromosomes in aneuploid cells, but nothing striking.

@@

```{r}
filtered_cnv_df <- cnv_df |> 
  dplyr::filter(copykat.pred != "not.defined")

caret_df_list <- filtered_cnv_df |> 
  dplyr::mutate(copykat = ifelse(
    # use str_detect for test data
    copykat.pred == "diploid", "Normal", "Tumor"
  )) |> 
  # make tumor the positive class
  dplyr::mutate(
    copykat = forcats::fct_relevel(copykat, "Tumor"),
    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
  ) |> 
  split(cnv_df$reference_used)

caret_df_list |> 
  purrr::imap(\(df, ref_type){
    caret::confusionMatrix(
      table(
        df$marker_gene_classification, 
        df$copykat)
    )
  })
```

@@

```{r}
# calculate Jaccard similarity index for each reference type
jaccard_matrices <- caret_df_list |>
  purrr::map(\(df) {

    make_jaccard_matrix(
      df,
      "marker_gene_classification",
      "copykat.pred"
    )
  })
```

@@

- `CopyKAT` seems to do a good job in this Ewing sarcoma sample of identifying aneuploid and diploid cells. 
- There is some small evidence of copy number gains in Chr8 and Chr12 which is consistent with findings in Ewing sarcoma patients. 
- The annotations from `CopyKAT` mostly line up with annotations obtained using marker gene annotation. 
- Using a reference of normal cells doesn't change the overall results but does reduce the run time for `CopyKAT`.
